---
title: "bike_data_analysis_2020_2021"
author: "Calvin"
date: "02/06/2021"
output: html_document
---

# Ask

* The goal of the task is to increase conversion rate for casual riders into annual members.
* We want to investigate the question that might lead to our goal:
" How do annual members and casual riders use Cyclistic bikes differently"

# Prepare

* The data used is first-party data collected by Cyclistic. (i.e data is ROCCC)
* Now to check data integrity.

```{r load packages, message=FALSE}
library(tidyverse)
library(sqldf)
library(ggplot2)
library(scales)
```



```{r read data, message=FALSE}
apr_2021 <- read_csv("bike_data/202104-divvy-tripdata.csv")
mar_2021 <- read_csv("bike_data/202103-divvy-tripdata.csv")
feb_2021 <- read_csv("bike_data/202102-divvy-tripdata.csv")
jan_2021 <- read_csv("bike_data/202101-divvy-tripdata.csv")
dec_2020 <- read_csv("bike_data/202012-divvy-tripdata.csv")
nov_2020 <- read_csv("bike_data/202011-divvy-tripdata.csv")
oct_2020 <- read_csv("bike_data/202010-divvy-tripdata.csv")
sept_2020 <- read_csv("bike_data/202009-divvy-tripdata.csv")
aug_2020 <- read_csv("bike_data/202008-divvy-tripdata.csv")
july_2020 <- read_csv("bike_data/202007-divvy-tripdata.csv")
june_2020 <- read_csv("bike_data/202006-divvy-tripdata.csv")
may_2020 <- read_csv("bike_data/202005-divvy-tripdata.csv")
apr_2020 <- read_csv("bike_data/202004-divvy-tripdata.csv")

# check type of variable in all data
str(apr_2021)
str(mar_2021)
str(feb_2021)
str(jan_2021)
str(dec_2020)

# found station id is num type before nov_2020
str(nov_2020)
str(oct_2020)
str(sept_2020)
str(aug_2020)
str(july_2020) 
str(june_2020)
str(may_2020)
str(apr_2020)
```


# Process

* We want to clean data if contain any missing value and do some transformation for additional information.


``` {r combine data}
# change variable type
nov_2020 <- nov_2020 %>% mutate(start_station_id = as.character(start_station_id),
                    end_station_id = as.character(end_station_id))

oct_2020 <- oct_2020 %>% mutate(start_station_id = as.character(start_station_id),
                    end_station_id = as.character(end_station_id))

sept_2020 <-sept_2020 %>% mutate(start_station_id = as.character(start_station_id),
                    end_station_id = as.character(end_station_id))

aug_2020 <-aug_2020 %>% mutate(start_station_id = as.character(start_station_id),
                    end_station_id = as.character(end_station_id))

july_2020 <- july_2020 %>% mutate(start_station_id = as.character(start_station_id),
                    end_station_id = as.character(end_station_id))

june_2020 <- june_2020 %>% mutate(start_station_id = as.character(start_station_id),
                    end_station_id = as.character(end_station_id))

may_2020 <- may_2020 %>% mutate(start_station_id = as.character(start_station_id),
                    end_station_id = as.character(end_station_id))

apr_2020 <- apr_2020 %>% mutate(start_station_id = as.character(start_station_id),
                    end_station_id = as.character(end_station_id))



# combine all the variables
all_bike_data <- bind_rows(apr_2021,mar_2021,feb_2021,jan_2021,dec_2020,nov_2020,
                           oct_2020,sept_2020,aug_2020,july_2020,june_2020,may_2020,apr_2020)
```

```{r cleaning and transforming}

# drop some irrelevant columns
all_bike_data <- all_bike_data %>% select(-c(start_lat, start_lng, end_lat, end_lng))

# check missing values

colSums(is.na(all_bike_data))

# There are some missing value in station id/name. But it is OK since we mainly want to investigate 
# difference of membership type.

# rename member_casual
all_bike_data <- all_bike_data %>% rename(membership = member_casual)

# Now add ride length by subtract ended_at - started_at
all_bike_data <- all_bike_data %>% mutate(ride_length = ended_at - started_at)

# We also want information of days of week
all_bike_data <- all_bike_data %>% mutate(day_of_week =  weekdays(as.Date(all_bike_data$started_at)))

# Remove negative length
all_bike_data <- all_bike_data %>% filter(ride_length >=0)

```
# Analyze

```{r Analyze by visualization}

# check average ride length for different membership
aggregate(all_bike_data$ride_length ~ all_bike_data$membership, FUN = mean)
aggregate(all_bike_data$ride_length ~ all_bike_data$membership, FUN = median)

all_bike_data %>% 
  group_by(membership) %>%
  summarize(Average_ride_length = mean(ride_length)) %>%
  ggplot(., aes(x=membership, y= Average_ride_length, fill = membership)) +
  geom_bar(stat='identity') +
  labs(title="Average ride time for Casual vs Member", y = "Average ride length in second") +
  scale_y_continuous(label = number)
```

```{r message=FALSE}

# use to correct order in graph
day_level = c('Monday','Tuesday',"Wednesday", "Thursday", "Friday", "Saturday", "Sunday")


all_bike_data %>% 
  group_by(membership, day_of_week) %>%
  summarize(Average_ride_length = mean(ride_length), .groups = 'drop') %>%
  ggplot(., aes(x=factor(day_of_week, level = day_level ), y= Average_ride_length, fill = membership)) +
  geom_bar(stat='identity',position = "dodge") +
  labs(title="Average ride time for Casual vs Member in day of week", y = "Average ride length in second", x="day of the week")
```
```{r}

# check count for each day of week

# use to correct order in graph
day_level = c('Monday','Tuesday',"Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

# check if day of week relate to ride length and membership
all_bike_data %>% 
  group_by(membership, day_of_week) %>%
  summarize(number_rides= n(), .groups="drop") %>%
  arrange(factor(day_of_week, level = day_level))

all_bike_data %>% 
  ggplot(., aes(x=factor(day_of_week, level = day_level ), fill = membership)) +
  geom_bar(position = "dodge") +
  labs(title="Number of rides for Casual vs Member", y = "Number of rides", x="day of the week") +
  scale_y_continuous(labels = number)
```
# Key finding

* Average casual ride time is 2.7 times more than average member ride time.
* Number of Casual ride are much higher in Friday and weekends than weekdays.
